Making Sequential Programs Deterministic

### 2019-03-23 design.tex

\section{Design}\label{design}
\subsection{Type checking background}\label{type-checking}
The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
The hierarchy of types in this lattice defines subtyping relationships among them.
In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.15]{lattice}
    \end{center}
    \caption{Example type hierarchy}
    \label{fig-example-lattice}
\end{figure}
In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
\begin{figure}
    \begin{verbatim}
    @A int x; @B int y; @C int z;
    x = y;    // Because types @A and @B are incomparable.
    z = x;    // Because @C is a subtype of @A.
    \end{verbatim}
\caption{Example: Invalid assingment}
\label{code-invalid1}
\end{figure}

Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
checker would report an error at compile time.

%Method overriding and example
For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
in \<class Y>.
\begin{figure}
    \begin{verbatim}
    class X {
        @B int foo(@C int param) {}
    }
    class Y extends X {
        @C int foo(@B int param) {}
    }
    \end{verbatim}
    \caption{Example: Invalid Method override}
    \label{code-invalid2}
\end{figure}

The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).

%collections type parameters - invariant
Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.

%covariant array types
Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
declared as \<@B int @A[] y> would type check without any warning.

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
\<@Top> and \<@Bottom> types for brevity):
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@OrderNonDet> indicates that
    a collection (i.e any class that is a subtype of java.util.Collection, java.util.Iterator, or java.util.Map) or an array will have the same elements in every execution, but in a
    possibly different order.  \<@OrderNonDet> may only be written on
    collections and arrays.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

\subsection{Determinism}\label{determinism}

@@

\subsection{Additional typing rules for collection elements}\label{collection-rules}
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

>>>>

\section{Design}\label{design}

%% This section is too basic, and therefore is boring because most readers
%% will already know it.  It also delays the interesting parts of the paper
%% for too long.  We can introduce concepts as they are needed in the rest
%% of the paper.
% \subsection{Type checking background}\label{type-checking}
% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
% The hierarchy of types in this lattice defines subtyping relationships among them.
% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
% \begin{figure}
%     \begin{center}
%         \includegraphics[scale=0.15]{lattice}
%     \end{center}
%     \caption{Example type hierarchy}
%     \label{fig-example-lattice}
% \end{figure}
% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
% \begin{figure}
%     \begin{verbatim}
%     @A int x; @B int y; @C int z;
%     x = y;    // Because types @A and @B are incomparable.
%     z = x;    // Because @C is a subtype of @A.
%     \end{verbatim}
% \caption{Example: Invalid assingment}
% \label{code-invalid1}
% \end{figure}
% 
% %Method overriding and example
% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
% in \<class Y>.
% \begin{figure}
%     \begin{verbatim}
%     class X {
%         @B int foo(@C int param) {}
%     }
%     class Y extends X {
%         @C int foo(@B int param) {}
%     }
%     \end{verbatim}
%     \caption{Example: Invalid Method override}
%     \label{code-invalid2}
% \end{figure}
% 
% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
% 
% %collections type parameters - invariant
% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
% 
% %covariant array types
% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
% declared as \<@B int @A[] y> would type check without any warning.

\todo{Need some background around here about what a type qualifier is,
  how it is represented in Java, and how to read a type \<@Q BaseType>.}

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@OrderNonDet> indicates that
    a collection or an array will have the same elements in every execution, but in a
    possibly different order.  \<@OrderNonDet> may only be written on
    collections and arrays.
    A collection is any subtype of \<java.util.Collection>,
    \<java.util.Iterator>, or \<java.util.Map>, including user-defined types.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

\subsection{Polymorphism}\label{polymorphism}

@@

\subsection{Additional typing rules for collection elements}\label{collection-rules}

\todo{It is very important to express the whole section as formal type
  jugdments, or we will get laughed
  out of the room as amateurs, during the program committee meeting.  Also,
  the formalism will help make the ideas precise.

  Type well-formedness rules:\todo{format nicely in \LaTeX}

  antecedent: $\Gamma : \kappa_e <: \kappa_c$

  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
  \codeid{Object}\rangle$
  }

The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:
